A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-08-29 paper/project.tex
Edits to figures for 1 and 2 rods, some intro work

Classical density-functional theory (DFT) provides an efficient way to evaluate the
free energy using the density of the fluid. The free energy is a functional of the 
density function and is independent of any external potentials\cite{ebner1976}. DFT 
can be used with a variety of external potentials, from the simple interaction with a 
hydrophobic wall to the complex strong interaction with hydrophilic solutes.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water.

@@

Figure \ref{fig:densityresolution} shows the density profiles for three different
values of the lattice spacing resolution. The density oscillations outside a single
rod of radius 0.5 nm look nearly identical for all resolutions except for the first peak. 
This contact density is important to get correct as it \fixme{relates to energy}. The 
contact density calculated using the medium resolution is within 3.5\% of the value at
high resolution, while the low resolution contact density is only within 5.1\%. The total
free energy calculations compare much more favorably. The medium resolution is within 
0.5\% and the low resolution is within 1.0\% of the high resolution free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profile for two hydrophobic rods.}
\label{fig:density-rods}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two identical hydrophobic rods with diameters
of 1 nm. }
\label{fig:rods-energy-vs-distance}
\end{figure}

>>>>

Classical density-functional theory (DFT) provides an efficient way to evaluate the
free energy using the density of the fluid. The free energy is a functional of the 
density function and is independent of any external potentials\cite{ebner1976}. 
Mermin\cite{Mermin} gives a theorem extending that of Hohenberg and Kohn\cite{HK} 
to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r) d^3r\right\}
\end{equation}
This is the basis for classical DFT, which can be used with a variety of external potentials, 
from the simple interaction with a 
hydrophobic wall to the complex strong interaction with hydrophilic solutes.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure \ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3 nm, 0.5 nm, and 1 nm. There are the
expected oscillations outside the rod and as the distance away from the rod increases
the density approaches that of the bulk saturated liquid density. The small radius
rod has comparatively larger oscillations and a higher contact density.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-single-rod}
\end{center}
\caption{ Density profiles for a single rod of different radii. The dotted line
represents the saturated liquid density.  }
\label{fig:density-single-rod}
\end{figure}

@@

Figure \ref{fig:densityresolution} shows the density profiles for three different
values of the lattice spacing resolution. The density oscillations around a single
rod of radius 0.5 nm look nearly identical for all resolutions except for the first peak. 
This contact density is important to get correct as it \fixme{relates to energy}. The 
contact density calculated using the medium resolution is within less than 4\% of the value at
high resolution, while the low resolution contact density is off by about 5\%. There is some
uncertainty in the value of contact density due to the different lattice spacings. The total
free energy calculations compare much more favorably. The medium resolution is within 
0.5\% and the low resolution is within 1\% of the high resolution free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profile for two hydrophobic rods (diameter 1.0 nm) at a 
separation of 0.8 nm. The scale on the right is density in g/mL.}
\label{fig:density-rods}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two identical hydrophobic rods with diameters
of 1.0 nm. }
\label{fig:rods-energy-vs-distance}
\end{figure}
